A general and efficient representation of Ancestral Recombination Graphs

### 2022-08-03 paper.tex
ARG process appendix

In the 1990s, Griffiths and colleagues revisited the
coalescent with recombination from a different perspective,
in which the stochastic process itself is encoded as a graph
(with particular properties; see section XXX for details),
and showed how the local trees generated by Hudson's algorithm
were embedded within
it~\citep{griffiths1991two,ethier1990two,griffiths1996ancestral,griffiths1997ancestral}.
They referred to this graph,
representing both the stochastic process and
its random realisations, as the Ancestral Recombination Graph (ARG).
Although mathematically equivalent, it is
important to note that the Griffiths and Hudson formulations of
the coalescent with recombination are not identical;
in particular, a direct implementation of the ARG process
requires exponential time to simulate
(see Appendix XXX for details). However, ARGs provided a way
to describe recombinant ancestry as single object that
could be inferred in a way
that was missing from Hudson's treatments.

@@

The likelihood $p(D | G)$ is effectively always the distribution of
a Poisson process on the ancestral material along the edges of $G$ \citep[Eq.\ (2)]{mahmoudi2022bayesian}.
Hence it is typically straightforward to evaluate in principle, but choosing the
right data structure can still have a dramatic effect on the efficiency of evaluation
\citep{mahmoudi2022bayesian}. Especially in the Bayesian case, $p(G)$ is typically specified
as the sampling distribution of a stochastic process, such as the coalescent with
recombination. Two contemporary examples are
\cite{mahmoudi2022bayesian, guo2022recombination}.
Evaluating the distribution of the coalescent with recombination
(or related models) requires knowledge of the number of extant lineages and
number of links available for recombination (i.e.\ ones at which a recombination
would split ancestral material) at each time \citep[Eq.\ (3)]{mahmoudi2022bayesian}.
These require a data structure from which shared edges on different trees are easy to identify,
and which encodes recombination events and times explicitly, e.g.\ using nodes.

@@

\begin{figure}[ht]
\centering
% FIXME this is a quick nasty hack to make the figure a bit smaller and
% prevent flushing all the other figures to the end of the document.
\scalebox{0.5}{
\begin{tabular}{cc}
\begin{tikzpicture}
	\node [anchor=north west] at (-3.5,9.3) {\LARGE \textbf{A}};
	\draw (-0.4, -0.2) -- (0.4, -0.2) -- (0.4, -0.3) -- (-0.4, -0.3) -- (-0.4, -0.2);
	\draw (-0.1, -0.1) -- (-0.1, -0.4);
	\node [label=below:{\small $v$}] at (-0.1, -0.2) {};
	\draw (3.6, -0.2) -- (4.4, -0.2) -- (4.4, -0.3) -- (3.6, -0.3) -- (3.6, -0.2);
	\draw (5.6, -0.2) -- (6.4, -0.2) -- (6.4, -0.3) -- (5.6, -0.3) -- (5.6, -0.2);

@@

	% Dashed lines for start and end times
	\draw[dashed] (-2, 0) -- (11.1, 0);
	\node [label=left:{$t = 0$}] at (-2,0) {};
	\draw[dashed] (-2, 8) -- (11.1, 8);
	\node [label=left:{$t = T$}] at (-2,8) {};

@@

	% Gray dashed lines to visually separate holding times
	\draw[color=gray, dashed] (7.5, 1.2) -- (11.1, 1.2);
	\draw[color=gray, dashed] (7.5, 1.7) -- (11.1, 1.7);
	\draw[color=gray, dashed] (7.5, 2.4) -- (11.1, 2.4);
	\draw[color=gray, dashed] (7.5, 3.1) -- (11.1, 3.1);
	\draw[color=gray, dashed] (7.5, 4.5) -- (11.1, 4.5);
	\draw[color=gray, dashed] (7.5, 5.4) -- (11.1, 5.4);
	\draw[color=gray, dashed] (7.5, 6) -- (11.1, 6);
	\draw[color=gray, dashed] (7.5, 7.2) -- (11.1, 7.2);
\end{tikzpicture}
&
\begin{tikzpicture}
	\node [anchor=north west] at (-3.5,9.3) {\LARGE \textbf{B}};
	\draw (-0.4, -0.2) -- (0.4, -0.2) -- (0.4, -0.3) -- (-0.4, -0.3) -- (-0.4, -0.2);
	\draw (-0.1, -0.1) -- (-0.1, -0.4);
	\node [label=below:{\small $v$}] at (-0.1, -0.2) {};
	\draw (3.6, -0.2) -- (4.4, -0.2) -- (4.4, -0.3) -- (3.6, -0.3) -- (3.6, -0.2);
	\draw (5.6, -0.2) -- (6.4, -0.2) -- (6.4, -0.3) -- (5.6, -0.3) -- (5.6, -0.2);

@@

@@

	% Gray dashed lines to visually separate holding times
	\draw[color=gray, dashed] (7.5, 1.2) -- (11, 1.2);
	\draw[color=gray, dashed] (7.5, 1.7) -- (11, 1.7);
	\draw[color=gray, dashed] (7.5, 2.4) -- (11, 2.4);
	\draw[color=gray, dashed] (7.5, 3.1) -- (11, 3.1);
	\draw[color=gray, dashed] (7.5, 3.8) -- (11, 3.8);
	\draw[color=gray, dashed] (7.5, 4.5) -- (11, 4.5);
	\draw[color=gray, dashed] (7.5, 5.4) -- (11, 5.4);
	\draw[color=gray, dashed] (7.5, 6) -- (11, 6);
	\draw[color=gray, dashed] (7.5, 6.6) -- (11, 6.6);
	\draw[color=gray, dashed] (7.5, 7.2) -- (11, 7.2);
\end{tikzpicture}
\end{tabular}
}
\caption{(A)
A realisation of the graph traversed by Hudson's algorithm started from a
sample of three chromosomes of length $m$ at time $t = 0$, and
propagated until time $T$. The MRCA on the genetic interval $[v, w)$ is reached
at event \textsf{b}, while that on $[0, v)$ is reached at event \textsf{c}.
The non-ancestral segment $[v, w)$ above
A contributes to the rate of effective recombinations because it
is trapped between ancestral segments. The two columns titled CA and RE
are the respective rates of mergers and recombinations when
the recombination rate is $\rho$.
(B) A corresponding realisation of a big ARG, which augments Hudson's algorithm
by tracking nonancestral lineages. The result is a simpler state space and
dynamics, at the cost of extra nodes and edges, highlighted in red, which do
not affect the local tree at any site.}
\label{hudson_vs_bigARG}
\end{figure}

[JK: this is text roughly related to big vs little ARG taken from various
iterations]

In the Griffiths formulation, each edge in the graph corresponds to an extant
lineage and nodes are events in the process (the initial $n$ leaf nodes are
``sampling'' events). Common ancestor events occur at rate $\binom{k}{2}$ when there
are $k$ lineages present. We choose two lineages (edges) uniformly, and merge them
into a common ancestor lineage. Recombination events happen at a rate of
$k \rho$. Here, we choose a lineage (edge) uniformly, and a
breakpoint $0 < x < m$ uniformly on its genome. We terminate the edge at a
node, record the breakpoint and start two new edges from this node. The process
then continues until there is only one lineage left (the Grand Most Recent
Common Ancestor, GMRCA), which is guaranteed to
happen in finite time because of the linear vs quadratic rates of branching
and coalescing. Later, we will show how the graph structure and breakpoints
associated with
recombination nodes provides sufficient information to later recover the marginal
trees and calculate sampling probabilities. However, this process traces out
a larger graph (the ``big'' ARG) than the Hudson algorithm (the ``little'' ARG; see the ``\nameref{Ancestry_resolution}'' section),
and is consequently considerably less efficient \citep{wiuf1999ancestry}.

The state-space of the Griffiths process is much simpler than Hudson's algorithm,
which greatly facilitates mathematical reasoning. This simplicity comes at a
substantial cost, however, if we wish to use it as a practical means of
simulating recombinant ancestry. The big ARG is vast, and any realisation
for even moderate levels of recombination is far too large to be of practical
use. The number of events in the big ARG all the way back to the GMRCA
is $O(e^\rho)$~\citep{griffiths1997ancestral}, whereas the number
of events required to simulate the little ARG is
$O(\rho^2)$~\citep{hein2004gene,baumdicker2021efficient}.
This disparity in the number of events in the two formulations is
because the majority of the events that occur in the big ARG do
not affect the genetic ancestry of the sample in any way. Recombination
events that occur outside of ancestral material do not have any bearing
on the ancestry of the sample, and so the structure is hugely redundant.
As~\cite{wiuf1999recombination} note,
``an `ancestral' sequence in the birth and death process
need not have any genetic material in common with a
sequence descended from it.''

@@

% % This is a discussion of Hudson's alg left over from earlier version. We
% % might wish to adapt some of this when defining the Little ARG.

% Each lineage consists of a list of
% disjoint ancestry segments $(\ell, r, a)$, where
% $[\ell, r)$ is a half-closed genomic interval and $a$ is an integer
% tracking the number of samples to which the lineage is ancestral over that interval.
% (We also usually track the tree node associated with each segment, but
% that is not important for our purposes here so we omit it.)
% If we have $n$ samples and a genome of length $m$, the process begins with $n$ lineages
% of the form $\{(0, m, 1)\}$. The process then works backwards in time from
% the present day as a series of random common ancestor or recombination events.
% Recombination events occur at a rate determined by the amount of ancestral material and
% the way in which it is distributed along a chromosome.
% Let $L$ be the set of ancestors at a given time $t$. Recombination events
% happen at rate $\rho \nu / (m - 1)$ where
% \[
% \nu = \sum_{x \in L}\left( \max_{(\ell, r, a) \in x}r
%     - \min_{(\ell, r, a) \in x}\ell - 1 \right)
% \]
% is the number of available `links' that may be broken. Thus, the rate of
% recombination is determined by the left- and right-most extent of the
% ancestral material carried by each lineage. At a recombination
% event we choose one of these links uniformly and break it. Given a lineage
% $x = [(\ell_j, r_j, a_j)]$ and a breakpoint $k$, we have two lineages
% $x_1$ and $x_2$ such that FILL IN DETAILS

% When $k = |L|$ lineages are present, common ancestor events
% occur at rate $\binom{k}{2}$. In a common ancestor event, two lineages
% are chosen uniformly at random and their ancestry segments merged.
% If we have overlapping intervals of ancestry from the two lineages,
% say, $(\ell, r, a_1)$ and $(\ell, r, a_2)$, a
% \emph{coalescence} occurs and ancestor represented by the current event
% will be present as a node (at least) in the marginal trees covering
% the interval $[\ell, r)$. The result of this coalescence is a segment
% $(\ell, r, a_1 + a_2)$, and if $a_1 + a_2 < n$ it is included in the
% ancestry for the new lineage. Otherwise, if $a_1 + a_2 = n$ we know that
% we have found the most recent common ancestor of all samples in
% the interval $[\ell, r)$ and so we do not need to simulate its history any further.
% Nonoverlapping intervals of ancestry from the two lineages are included
% in the resulting lineage without changes. Eventually, as the process continues,
% we find resultant lineages in which all segments have fully coalescenced,
% and so the number of extant lineages gradually dwindles down to zero.

>>>>

In the 1990s, Griffiths and colleagues revisited the
coalescent with recombination from a different perspective,
in which the stochastic process itself is encoded as a graph
(with particular properties; see section XXX for details),
and showed how the local trees generated by Hudson's algorithm
were embedded within
it~\citep{griffiths1991two,ethier1990two,griffiths1996ancestral,griffiths1997ancestral}.
They referred to this graph,
representing both the stochastic process and
its random realisations, as the Ancestral Recombination Graph (ARG).
Although mathematically equivalent, it is
important to note that the Griffiths and Hudson formulations of
the coalescent with recombination are not identical;
in particular, a direct implementation of the ARG process
requires exponential time to simulate
(see Appendix XXX for details). However, ARGs provided a way
to describe recombinant ancestry as single object that
could be inferred in a way
that was missing from Hudson's treatments.
An overview of the Hudson and Griffiths stochastic processes is provided in the appendix.

@@

The likelihood $p(D | G)$ is effectively always the distribution of
a Poisson process on the ancestral material along the edges of $G$ \citep[Eq.\ (2)]{mahmoudi2022bayesian}.
Hence it is typically straightforward to evaluate in principle, but choosing the
right data structure can still have a dramatic effect on the efficiency of evaluation
\citep{mahmoudi2022bayesian}. Especially in the Bayesian case, $p(G)$ is typically specified
as the sampling distribution of a stochastic process, such as the coalescent with
recombination. Two contemporary examples are
\cite{mahmoudi2022bayesian, guo2022recombination}.
Evaluating the distribution of the coalescent with recombination
(or related models) requires knowledge of the number of extant lineages and
number of links available for recombination (i.e.\ ones at which a recombination
would split ancestral material) at each time \citep[Eq.\ (3)]{mahmoudi2022bayesian}.
These require a data structure from which shared edges on different trees are easy to identify,
and which encodes recombination events and times explicitly, e.g.\ using nodes.
We will provide a more detailed discussion on sampling probability evaluation in the appendix.

@@

Here we review the two predominant stochastic processes which construct ARGs:
the ``big" ARG process of \cite{griffiths1997ancestral}, and the ``little" ARG process of
 \cite{hudson1983properties}. The big ARG process is mathematically simpler
 but is computationally intractable due to generating a vast number of ancestors
 which contribute no genetic material to the initial sample.
The little ARG process avoids non-genetic ancestors at the cost of more complex
dynamics and state space. We also demonstrate that evaluating the sampling probability
of either process---a key quantity in many statistical approaches---requires that the
gARG (or eARG) data structure be interpreted in a model-specific way.

A generic state of the little ARG process consists of a finite collection of lineages $L$,
each of which is a list of disjoint ancestry segments $(\ell, r, a)$, where
$[\ell, r)$ is a half-closed genomic interval and $a$ is an integer
tracking the number of samples to which the lineage is ancestral over that interval.
We also usually track the node associated with each segment, but
that is not important for our purposes here so we omit it to lighten notation.
The initial condition for a sample of $n$ genomes of length $m$ consists of $n$ lineages 
of the form $\{(0, m, 1)\}$. The process traverses a series of common ancestor and 
recombination events backwards in time.
Recombination events happen at rate $\rho \nu / (m - 1)$,
where $\rho \geq 0$ is a per-genome recombination rate and
 \[
 \nu = \sum_{x \in L}\left( \max_{(\ell, r, a) \in x}r
     - \min_{(\ell, r, a) \in x}\ell - 1 \right)
 \]
 is the number of available ``links" surrounded by ancestral material.
 At a recombination event we choose one of these links uniformly and break it,
 replacing the original lineage in $L$ with two new lineages containing the ancestral material
 to the left and right of the break point, respectively.

Common ancestor events occur at rate $\binom{|L|}{2}$.
In a common ancestor event, two uniformly sampled lineages have their segments
merged into a single ancestor lineage, which is added to $L$.
If the lineages have overlapping intervals of ancestry,
say, $(\ell, r, a_1)$ and $(\ell, r, a_2)$, a
\emph{coalescence} occurs. The result is a segment
$(\ell, r, a_1 + a_2)$, and if $a_1 + a_2 < n$ it is included in the
ancestor lineage. Otherwise, if $a_1 + a_2 = n$, we have found
the most recent common ancestor of all samples in the interval$[\ell, r)$
and do not need to simulate its history any further.
Non-overlapping intervals from the two lineages are included
 in the ancestor lineage without changes. Eventually,
we find resultant lineages in which all segments have fully coalesced,
and so the number of extant lineages gradually falls to zero.

In the Griffiths formulation (the big ARG process), each edge in the graph corresponds to an extant
lineage and nodes are events in the process. The $n$ initial leaf nodes are
sampling events. Common ancestor events occur at rate $\binom{|L|}{2}$.
When a common ancestor event happens, two uniformly chosen lineages
merge into a common ancestor lineage.
Recombination events happen at rate $|L| \rho$. Here, we choose a lineage (i.e.\ edge) uniformly,
and a breakpoint $0 < x < m$ uniformly on its genome. We terminate the edge at a
node, record the breakpoint, and start two new edges from this node. The process
then continues until there is only one lineage left (the Grand Most Recent
Common Ancestor, GMRCA), which is guaranteed to
happen in finite time because of the linear rate of coalescing vs.\ quadratic rate of branching.

The state-space of the big ARG process is much simpler than that of the little ARG process,
which greatly facilitates mathematical reasoning. This simplicity comes at a
substantial cost, however, if we wish to use it as a practical means of
simulating recombinant ancestries. 
The number of events in the big ARG all the way back to the GMRCA
is $O(e^\rho)$~\citep{griffiths1997ancestral}, whereas the number
of events required to simulate the little ARG is
$O(\rho^2)$~\citep{hein2004gene,baumdicker2021efficient}.
This disparity arises because the majority of the events in the big ARG are 
recombination events which occur outside of ancestral material,
and this do not have any bearing on the ancestry of the initial sample.
Because we don't keep track of the distribution of ancestral material during the process,
we generate a vastly larger graph.

Figure \ref{hudson_vs_bigARG} illustrates the more complex state space
of the little ARG process, as well as the extra events which occur in the big ARG process.
Moreover, it depicts the rates of common ancestors and recombination events in each
interval of time of the realisations.
In order to evaluate these rates, and hence the sampling probability
(see e.g.\ \cite[Equation (3)]{mahmoudi2022bayesian}),
it is necessary to know the number of lineages and number of extant links
available for recombination in each time interval.
This information cannot be recovered from the gARG encoding depicted in
e.g.\ Figure \ref{fig-ancestry-resolution}.
For example, it is clear that a recombination takes place between nodes \textsf{b} and
\textsf{d} as well as \textsf{e}, but the exact time of the event is ambiguous,
and thus so is the number of lineages during the time interval.
Conventions can be introduced to resolve such ambiguities;
for the likelihood-based inference algorithms for the coalescent with recombination,
the two parent nodes are typically created at the time of the recombination event.
See the appendix of \cite{baumdicker2021efficient} for details of evaluating
coalescent with recombination likelihoods using this convention.
This is also the interpretation depicted in
Figure \ref{hudson_vs_bigARG}, but it means that the two edges above node \textsf{b}
in Figure \ref{fig-ancestry-resolution} should correspond to only one lineage,
along which all $m-1$ links are available for recombination.
The lineage then splits into two at the time of nodes \textsf{d} and \textsf{e}.
Nodes \textsf{k}, \textsf{l}, and \textsf{m} in Figure \ref{fig-ancestry-resolution}
demonstrate that the same issue can affect the number of links available for recombination:
without an external convention, the exact time at which the trapped ancestral material on
node \textsf{k} ceases to be available for an effective recombination in the little ARG process.

\begin{figure}[ht]
\centering
% FIXME this is a quick nasty hack to make the figure a bit smaller and
% prevent flushing all the other figures to the end of the document.
\scalebox{0.54}{
\begin{tabular}{cc}
\begin{tikzpicture}
	\node [anchor=north west] at (-3.5,9.3) {\LARGE \textbf{A}};
	\draw (-0.4, -0.2) -- (0.4, -0.2) -- (0.4, -0.3) -- (-0.4, -0.3) -- (-0.4, -0.2);
	\draw (-0.1, -0.1) -- (-0.1, -0.4);
	\node [label=below:{\small $v$}] at (-0.1, -0.2) {};
	\draw (3.6, -0.2) -- (4.4, -0.2) -- (4.4, -0.3) -- (3.6, -0.3) -- (3.6, -0.2);
	\draw (5.6, -0.2) -- (6.4, -0.2) -- (6.4, -0.3) -- (5.6, -0.3) -- (5.6, -0.2);

@@

@@

	% Gray dashed lines to visually separate holding times
	\draw[color=gray, dashed] (7.5, 1.2) -- (11, 1.2);
	\draw[color=gray, dashed] (7.5, 1.7) -- (11, 1.7);
	\draw[color=gray, dashed] (7.5, 2.4) -- (11, 2.4);
	\draw[color=gray, dashed] (7.5, 3.1) -- (11, 3.1);
	\draw[color=gray, dashed] (7.5, 4.5) -- (11, 4.5);
	\draw[color=gray, dashed] (7.5, 5.4) -- (11, 5.4);
	\draw[color=gray, dashed] (7.5, 6) -- (11, 6);
	\draw[color=gray, dashed] (7.5, 7.2) -- (11, 7.2);
\end{tikzpicture}
&
\begin{tikzpicture}
	\node [anchor=north west] at (-3.5,9.3) {\LARGE \textbf{B}};
	\draw (-0.4, -0.2) -- (0.4, -0.2) -- (0.4, -0.3) -- (-0.4, -0.3) -- (-0.4, -0.2);
	\draw (-0.1, -0.1) -- (-0.1, -0.4);
	\node [label=below:{\small $v$}] at (-0.1, -0.2) {};
	\draw (3.6, -0.2) -- (4.4, -0.2) -- (4.4, -0.3) -- (3.6, -0.3) -- (3.6, -0.2);
	\draw (5.6, -0.2) -- (6.4, -0.2) -- (6.4, -0.3) -- (5.6, -0.3) -- (5.6, -0.2);

@@

	% Dashed lines for start and end times
	\draw[dashed] (-2, 0) -- (9.1, 0);
	\node [label=left:{$t = 0$}] at (-2,0) {};
	\draw[dashed] (-2, 8) -- (9.1, 8);
	\node [label=left:{$t = T$}] at (-2,8) {};

@@

	% Gray dashed lines to visually separate holding times
	\draw[color=gray, dashed] (7.5, 1.2) -- (9.1, 1.2);
	\draw[color=gray, dashed] (7.5, 1.7) -- (9.1, 1.7);
	\draw[color=gray, dashed] (7.5, 2.4) -- (9.1, 2.4);
	\draw[color=gray, dashed] (7.5, 3.1) -- (9.1, 3.1);
	\draw[color=gray, dashed] (7.5, 3.8) -- (9.1, 3.8);
	\draw[color=gray, dashed] (7.5, 4.5) -- (9.1, 4.5);
	\draw[color=gray, dashed] (7.5, 5.4) -- (9.1, 5.4);
	\draw[color=gray, dashed] (7.5, 6) -- (9.1, 6);
	\draw[color=gray, dashed] (7.5, 6.6) -- (9.1, 6.6);
	\draw[color=gray, dashed] (7.5, 7.2) -- (9.1, 7.2);
\end{tikzpicture}
\end{tabular}
}
\caption{(A)
A realisation of the graph traversed by Hudson's algorithm started from a
sample of three chromosomes of length $m$ at time $t = 0$, and
propagated until time $T$. The MRCA on the genetic interval $[v, w)$ is reached
at event \textsf{b}, while that on $[0, v)$ is reached at event \textsf{c}.
The non-ancestral segment $[v, w)$ above
A contributes to the rate of effective recombinations because it
is trapped between ancestral segments. The two columns titled CA and RE
are the respective rates of mergers and recombinations when
the recombination rate is $\rho$.
(B) A corresponding realisation of a big ARG, which augments Hudson's algorithm
by tracking nonancestral lineages. The result is a simpler state space and
dynamics, at the cost of extra nodes and edges, highlighted in red, which do
not affect the local tree at any site.}
\label{hudson_vs_bigARG}
\end{figure}

@@
